Progress Report

### 2023-09-19 Docs/ProgressReport.Rmd
data application

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
theme_set(theme_minimal())

df <- read_rds(here("Data/nhanes_bi.rds"))
load(here("Data/ApplOutput_fGFPCA.RData"))

@@

# NHANES data application

@@

```{r}
library(glmmTMB)
# re-fit local GLMM and check numeric issues
try_fit_lst<-list()
t1=Sys.time()
for(i in seq_along(df_bin_lst)){
  # this_glm <- tryCatch({glmer(Y ~ 1 + (1|id), data = df_bin_lst[[i]], family = binomial, nAGQ=5)},
  #                      warning=function(w){NA})
  this_glm <- glmmTMB(Y ~ 1 + (1|id), data = df_bin_lst[[i]], family = binomial)
  try_fit_lst[[i]] <- this_glm

}
df_est_latent_tmb <- lapply(try_fit_lst, function(x){predict(x, type = "link")})
t2=Sys.time()
t_tmb <- t2-t1
# population mean
```

@@

```{r, fig.width=12, fig.height=4}
mean_tmb <- sapply(try_fit_lst, function(x){coef(summary((x)))$cond[1]})

ggarrange(
df_est_local_marg %>% bind_rows() %>%
  select(bin, eta_hat) %>% distinct(.) %>%
  ggplot(aes(x=bin, y=eta_hat))+
  geom_line()+
  labs(x="local GLM (marginal)", y = "latent mean")+
  ylim(-15, 1),

@@

>>>>

library(here)
library(tidyverse)
library(ggpubr)
library(refund)
library(lme4)
library(ROCR)
library(GLMMadaptive)
library(kableExtra)
theme_set(theme_minimal())

df <- read_rds(here("Data/nhanes_bi.rds"))
# load(here("Data/ApplOutput_fGFPCA.RData"))

@@

# NHANES data

@@

```{r, cache=TRUE}
library(glmmTMB)
# re-fit local GLMM and check numeric issues
try_fit_lst<-list()
t1=Sys.time()
for(i in seq_along(df_bin_lst)){
  # this_glm <- tryCatch({glmer(Y ~ 1 + (1|id), data = df_bin_lst[[i]], family = binomial, nAGQ=5)},
  #                      warning=function(w){NA})
  this_glm <- glmmTMB(Y ~ 1 + (1|id), data = df_bin_lst[[i]], family = binomial)
  try_fit_lst[[i]] <- this_glm

}
df_est_latent_tmb <- lapply(try_fit_lst, function(x){predict(x, type = "link")})
t2=Sys.time()
t_tmb <- t2-t1
# population mean
mean_tmb <- sapply(try_fit_lst, function(x){coef(summary((x)))$cond[1]})
```

@@

```{r, fig.width=12, fig.height=4}
ggarrange(
df_est_local_marg %>% bind_rows() %>%
  select(bin, eta_hat) %>% distinct(.) %>%
  ggplot(aes(x=bin, y=eta_hat))+
  geom_line()+
  labs(x="local GLM (marginal)", y = "latent mean")+
  ylim(-15, 1),

@@

# NHANES data application output

## fGFPCA

- 80% (7017) subjects for model fitting, 20% (1753) subjects for out-of-sample prediction
- 3.3 minutes spent on model fitting, 20 minutes spent on prediction (about 0.01 per subject)
- Three subjects had failed approximation, all happened with the shortest observed track

```{r}
load(here("Data/ApplOutput_fGFPCA.RData"))
```

- Example of out-of-sample prediction

```{r message=FALSE, warning=FALSE}
rand_id2 <- sample(unique(df_test$id), 4)
df %>% 
  filter(id %in% rand_id2) %>%
  left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020)) %>%
  mutate_at(vars(pred_t540, pred_t780, pred_t1020), function(x)exp(x)/(1+exp(x))) %>% 
  ggplot()+
    geom_line(aes(x=bin, y = pred_t540, col = "9am"), linetype = "dashed")+
    geom_line(aes(x=bin, y = pred_t780, col = "1pm"),linetype = "dashed")+
    geom_line(aes(x=bin, y = pred_t1020, col = "5pm"),linetype = "dashed")+
    geom_point(aes(x=bin, y = Y, col = "Outcome"), size = 0.2)+
    facet_wrap(~id)+
    labs(x = "Time", y = "Estimated latent function (probablity scale)")
```

- Subjects with numeric issue

```{r}
df %>% filter(id %in% skip_id) %>% 
  ggplot()+
  geom_point(aes(x=sind, y=Y), size = 0.2)+
  facet_wrap(~id)+
  geom_vline(xintercept = c(540))
```

- Calculated AUC

Since we are assuming value of latent function is constant within a bin, why can't we just do the same when calculating AUC without interpolation?

Remove the three subjects with failed approximation

```{r}
lst_auc <- df %>% 
  filter(id %in% unique(df_test$id) & !id %in% skip_id ) %>%
  left_join(df_test %>% select(id, bin, pred_t540, pred_t780, pred_t1020), by = c("id", "bin")) %>% 
  mutate(window = cut(sind, breaks = c(0, 540, 780, 1020, 1260, 1440), 
                      labels = c("0-9am", "9am-1pm", "1pm-5pm", "5pm-9pm","9pm-12pm"), 
                      include.lowest = T))
lst_auc <- split(lst_auc, f=lst_auc$window)

# up to 9am
auc_t540 <- lst_auc[2:4] %>% 
  lapply(function(x)performance(prediction(x$pred_t540, x$Y), measure = "auc"))
auc_t540 <- lapply(auc_t540, function(x){x@y.values[[1]]}) %>% unlist()

# up to 1pm
auc_t780 <- lst_auc[3:4] %>% 
  lapply(function(x)performance(prediction(x$pred_t780, x$Y), measure = "auc"))
auc_t780 <- lapply(auc_t780, function(x){x@y.values[[1]]}) %>% unlist()

# up to 5pm
auc_t1020 <- lst_auc[4] %>% 
  lapply(function(x)performance(prediction(x$pred_t1020, x$Y), measure = "auc"))
auc_t1020 <- lapply(auc_t1020, function(x){x@y.values[[1]]}) %>% unlist()
```

## GLMMadaptvie

- when fitting model with **random intercept + slope** on the same full training set: Error: vector memory exhausted (limit reached?)
- when fitting model with **only random intercept** on the same full training set: Error regarding a large coefficient value. One fix is to re-scale covariates. So I re-scaled minute index by dividing them by J, so the range is within (0, 1], and tried again. This fitting procedure took 20.72 minutes to finish
- Out-of-sample prediction took 12.18 minutes. About 0.007 minute per subject.

```{r}
load(here("Data/ApplOutput_GLMMadaptive.RData"))
```

```{r}
# up to 540
auc_t540_adglmm <- adglmm_pred_t540$newdata2 %>%
  mutate(sind=sind*J) %>%
  mutate(window = cut(sind, breaks = c(540, 780, 1020, 1260, 1440), 
                      labels = c("9am-1pm", "1pm-5pm", "5pm-9pm","9pm-12pm"), 
                      include.lowest = T))
auc_t540_adglmm <- split(auc_t540_adglmm, f=auc_t540_adglmm$window)
auc_t540_adglmm <- auc_t540_adglmm[1:3] %>% 
  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
auc_t540_adglmm <- lapply(auc_t540_adglmm, function(x){x@y.values[[1]]}) %>% unlist()

# up to 780
auc_t780_adglmm <- adglmm_pred_t780$newdata2 %>%
  mutate(sind=sind*J) %>%
  mutate(window = cut(sind, breaks = c(780, 1020, 1260, 1440), 
                      labels = c("1pm-5pm", "5pm-9pm","9pm-12pm"), 
                      include.lowest = T))
auc_t780_adglmm <- split(auc_t780_adglmm, f=auc_t780_adglmm$window)
auc_t780_adglmm <- auc_t780_adglmm[1:2] %>% 
  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
auc_t780_adglmm <- lapply(auc_t780_adglmm, function(x){x@y.values[[1]]}) %>% unlist()

# up to 1020
auc_t1020_adglmm <- adglmm_pred_t1020$newdata2 %>%
  mutate(sind=sind*J) %>%
  mutate(window = cut(sind, breaks = c(1020, 1260, 1440), 
                      labels = c("5pm-9pm","9pm-12pm"), 
                      include.lowest = T))
auc_t1020_adglmm <- split(auc_t1020_adglmm, f=auc_t1020_adglmm$window)
auc_t1020_adglmm <- auc_t1020_adglmm[1] %>% 
  lapply(function(x)performance(prediction(x$pred, x$Y), measure = "auc"))
auc_t1020_adglmm <- lapply(auc_t1020_adglmm, function(x){x@y.values[[1]]}) %>% unlist()

## Compare fGFPCA and GLMMadaptive

```{r}
options(knitr.kable.NA = '')
tb_auc <- data.frame(auc_t540, 
           c(NA, auc_t780), 
           c(NA, NA, auc_t1020), 
           auc_t540_adglmm,
           c(NA, auc_t780_adglmm), 
           c(NA, NA, auc_t1020_adglmm))

colnames(tb_auc) <- rep(c("9am", "1pm", "5pm"), 2)

tb_auc %>%
  kable(digit = 2) %>%
  kable_styling(full_width = F) %>% 
  add_header_above(c("Time spent on model fitting" = 1, "3.3" = 3, "20.72" = 3), bold = F) %>%
  add_header_above(c("Time spent on prediction" = 1, "20.0" = 3, "12.2" = 3), bold = F) %>% 
  add_header_above(c(" "=1, "fGPFCA"=3, "GLMMadaptive"=3))
```

- fGFPCA almost always have higher AUC. It at least is as good as GLMMadaptive.
- Computationally, fGFPCA is about 10 minute faster (33%).

## Discussion

- What would be the reference method to compare our performance to? Original GLMMadaptvie perhaps is feasible on teh entire dataset. But what should we use if we wanna do sub-sample comparison? Should we even do that for this data application (or simulation alone)? 
- There are still three folks with numeric problem! Shall we extend the observation track even more? I suspect that may cause still the same issue just on different subjects.
